---
title: "Impact of Operational Intensity on Transportation Safety"
subtitle: "ECON 370 Final Project"
author: "Vikram Iyengar"
institute: "University of North Carolina at Chapel Hill"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      eforeInit: "libs/cols_macro.js"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---
```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#4B9CD3",
  header_font_google = google_font("Fira Sans","600"),
  text_font_google   = google_font("Fira Sans", "300", "300i"),
  base_font_size = "20px",
  text_font_size = "0.9rem",
  code_inline_font_size = "0.9em",
  code_font_size = "0.7rem",
  header_h1_font_size = "2.0rem",
  header_h2_font_size = "1.75rem",
  header_h3_font_size = "1.25rem",
  code_font_google   = google_font("Fira Code"),
  inverse_link_color = "#13294B",
  link_color = "#4B9CD3"
)
```
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
library(fontawesome)
opts_chunk$set(
  fig.align="center", fig.width=6, fig.height=3.5, 
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=T#, echo=F, warning=F, message=F
  )
```

# What Are We Studying

--

## Research Question: 
How do different factors of operational intensity correlate with overall
transportation safety?

--

## Motivation: 
Why do we want to study this?
- 

---
# Data
### Transportation Ridership Data:
**Transportation Services Index and Seasonally-Adjusted Transportation Data**
- Created by the U.S. Department of Transportation (DOT), Bureau of Transportation Statistics (BTS).
- Measures the movement of freight and passengers. 
- Note: Data is seasonally adjusted
  - Allows for measurement of real monthly changes; short and long term patterns of growth or decline; and turning points.

### Transportation Safety and Event Data:
**Modal Service data and Safety & Security (S&S) public transit time series data**
- Created by the U.S. Department of Transportation (DOT), (Federal Transit Administration).
- Measures statistics about transportation in regards to major and non-major safety events
```{r, include=F}
transport_data = read.csv("/Users/vikram/Documents/ECON 370/Final Project/Data/Transportation_Data.csv")
event_data = read.csv("/Users/vikram/Documents/ECON 370/Final Project/Data/Time_Series_Event_Data.csv")
```
---
# Key Variables
```{r, echo=FALSE}
data = list(c(1,2,3,4), c(1,5,3,2))
knitr::kable(data, caption = "Tables with kable")
```

--

What can we do here?

--

You are recommended to use the [RStudio IDE](https://www.rstudio.com/products/rstudio/), but you do not have to.

- Create a new R Markdown document from the menu `File -> New File -> R Markdown -> From Template -> Ninja Presentation`;<sup>1</sup>

--

- Click the `Knit` button to compile it;

--

- or use the [RStudio Addin](https://rstudio.github.io/rstudioaddins/)<sup>2</sup> "Infinite Moon Reader" to live preview the slides (every time you update and save the Rmd document, the slides will be automatically reloaded in RStudio Viewer.

.footnote[
[1] 中文用户请看[这份教程](https://slides.yihui.org/xaringan/zh-CN.html)

[2] See [#2](https://github.com/yihui/xaringan/issues/2) if you do not see the template or addin in RStudio.
]

---


```{r gen_pdf, include = FALSE, cache = FALSE, eval = TRUE}
infile = list.files(pattern = '.html')
#pagedown::chrome_print(input = infile, timeout = 100)
```